---
title: "Neuroinflammation effect on context discrimination memory: Data Import and Cleaning of raw training and testing behavioral data"
author: "G. Lewandowski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    fig_caption: yes
    float_toc: yes
    theme: journal
    toc: yes
  pdf_document:
    toc: yes
    
---

```{r setup, include=FALSE}
## get the type of document that is being rendered
out_type <- knitr::opts_knit$get("rmarkdown.pandoc.to")

knitr::opts_chunk$set(echo = TRUE)
```

## Document Introduction   

In this document the raw behavioral training and testing data for experiments 14x02 and 14x04 are imported. The imported data will be converted (if needed) to a long (tidy) data form.  The data will be compiled such that there is one behavioral training dataset and one behavioral testing dataset.   
Notably, for each dataset observation is synonymous with row and variable is synonymous with column.  Most frequently, variable will be used instead of column.   

## Contextual Discrimination Conditioning overview   

Contextual discrimination conditioning (CDC) memory was assessed by measuring freezing in two similar behavioral chambers.  There was a mild footshock in context A but not in context A'.  

For CDC training: on each training day the rats were place in each context for 3min 20s. For each min and for the final 20s interval the amount of freezing in seconds was recorded. The context order (AA' or A'A) was changed for each training day and was also recorded.  Rats were were trained to criterion as defined by **>=25% freezing in A and <=10% freezing in A' for 2 consecutive days**.  

For CDC testing: The testing data consist of one session in which the rats spent 3min 20s in context A' and then 5 min in context A.  Again freezing (in seconds) was recorded for each minute interval.   

Load R packages required for data import and cleaning:

```{r load libraries, echo=FALSE, results='hide'}
suppressMessages({
    library(rmarkdown)  ## render markdown docs (output)
    library(knitr)  ## knit markdown docs (output)
    library(readxl) ## read MS Excel worksheets into data tables
    library(pander)  ## generate tables in output
    library(broom) ## 'tidy()' summary output into a data frame
    library(stringr)  ## use substr() to extract info
    library(tidyverse)  ## data wrangling: dplyr, tidyr
})
```

## Data Import   
The behavioral data for 14x02 is contained in two worksheets within a Microsoft Excel workbook. Likewise, the behavioral data for 14x02 is contained in two worksheets within a Microsoft Excel workbook. In this step the data is imported from the Excel workbook into data tables.  Additionally, the context order for the training days (both 14x02 and 14x04) is stored in a worksheet in the 14x04 workbook.

```{r file path to behavioral data}
## path to 14x02 Excel workbook with experiment data (base R)
path_14x02 <- file.path("neuroinflam_CDC_raw_data",
    "14x02_raw_ behav_data.xlsx")

## path to 14x04 Excel workbook with experiment data
path_14x04 <- file.path("neuroinflam_CDC_raw_data",
    "14x04_raw_ behav_data.xlsx")

## path to the group assignments for 14x02 and 14x04
path_grps <- file.path("neuroinflam_CDC_raw_data",
    "group_assingments.xlsx")
```

The worksheets in the 14x02 Microsoft workbook are:  

```{r 14x02 worksheets}
## use the function excel_sheets() from the readxl package

## list the sheets in the 14x02 Excel workbook
excel_sheets(path_14x02)
```

The worksheets in the 14x04 Microsoft workbook are:  

```{r 14x04 worksheets}
## list the sheets in the 14x04 Excel workbook
excel_sheets(path_14x04)
```

For both 14x02 and 14x04 the "Freezing" worksheets contain the behavioral training data. The "Test Day" worksheets contain the behavioral testing data. And the "context_order" worksheet of the 14x04 workbook contains the context order for each training day.    

### Import 14x02 behavioral data   

```{r import 14x02 behav data}

## use the read_excel() function from the readxl package

## import 14x02 training data
train_14x02 <- read_excel(path = path_14x02,
                              sheet = "Freezing",
                              na = "")

## import 14x02 testing data
test_14x02 <- read_excel(path = path_14x02,
                              sheet = "Test Day",
                              na = "")

## save imported worksheets as RData objects
save(train_14x02, file = "original_14x02_train_dat.RData")
save(test_14x02, file = "original_14x02_test_dat.RData")
```

### Import 14x04 behavioral data   

```{r import 14x04 behav data}

## use the read_excel() function from the readxl package

## import 14x04 training data
train_14x04 <- read_excel(path = path_14x04,
                              sheet = "Freezing",
                              na = "")

## import 14x04 testing data
test_14x04 <- read_excel(path = path_14x04,
                              sheet = "Test Day",
                              na = "")

## import the context_order for the training days
context_order <- read_excel(path = path_14x04,
                              sheet = "context_order",
                              na = "")

## save imported worksheets as RData objects
save(train_14x04, file = "original_14x04_train_dat.RData")
save(test_14x04, file = "original_14x04_test_dat.RData")
save(context_order, file = "context_order_for_training.RData")
```

### Import behavioral group assingments   

```{r import grp assngmnts}
## import treatment group assignments
grp_assigns <- read_excel(path = path_grps,
                              sheet = "group_assignment",
                              na = "")

## save imported worksheet as RData object
save(grp_assigns, file = "grp_assigns.RData")

```

```{r rm path vars, echo=FALSE, results='hide'}
rm(path_14x02, path_14x04, path_grps)
```

## Initial Inspection of `grp_assigns`   
  1.  Overview of `grp_assigns`   
  
```{r glimpse grp}
## use glimpse() from the tibble package loaded with the tidyverse package
glimpse(grp_assigns)
```
The `grp_assigns` looks good and does not need any further processing.   

## Initial Training Data Inspection:   

Here we make an initial inspection of the imported behavioral training datasets and identify issues to resolve in order to generate the corresponding tidy datasets.    

### 14x02 Training Data Inspection   

  1.  Overview of the train_14x02 data:
  
```{r glimpse train_14x02}
## use glimpse() from the tibble package loaded with the tidyverse package

## overview of train_14x02
glimpse(train_14x02)
```
  
From this view of the 14x02 training data we see that:   
  *  There are `r dim(train_14x02)[1]` rows and `r dim(train_14x02)[2]` variables   
  *  The data appear to be in a long, tidy format   
  *  Rat identifiers are stored in the "Subject" variable   
  *  The raw freezing data of interest is in numeric variables with names containing the phrases "Min 1 sec", ".Min 2 sec", and "Min 3 sec"   
  *  The remaining numeric variables are values calculated from the raw data 
  *  Freezing data for context A is stored in variables with names starting with "S+"   
  *  Freezing data for context A' is stored in variables with names starting with "S-"   
  
  2.  Inspection of missing values:   
Are there rows in the 14x02 training data set that do not have **any** raw freezing data? If so, how many rows? To answer this question we look for rows in which the value for *all* the variables containing the raw freezing data are *NA* (missing).     

First, we can look at a vector of the number of missing raw data values for each observation.   

```{r train_14x02 NA by row }

rowSums(is.na(train_14x02[,c(3,5,7,11,13,15)]))
```

From this vector, we can see that starting with row (observation) 156, there are several observations missing 5 or 6 of the 6 values.  We can now find how many observations are missing >=5 values.  

```{r train_14x02 rows missing data}
## use select(), filter() and summarise() from dplyr

raw_dat <- train_14x02 %>%
    select(3,5,7,11,13,15) 

 num_NA_rows_14x02 <- raw_dat %>%
    filter(rowSums(is.na(raw_dat)) >= 6) %>%
    summarise(num_rows = n())
```

There are `r num_NA_rows_14x02[[1,1]]` rows (observations) without any raw freezing data, containing only NA values.  These observations will be removed during the Data Cleaning step.

### 14x04 Training Data Inspection   

  1. Overview of the train_14x04 data:
  
```{r glimpse train_14x04}
## use glimpse() from the tibble package loaded with the tidyverse package

## overview of train_14x04
glimpse(train_14x04)
```

From this view of the 14x04 training data we see that:   
  *  There are `r dim(train_14x04)[1]` rows and `r dim(train_14x04)[2]` variables   
  *  The data appear to be in exactly the same format as the 14x02 training data, with the same variables containing the same types of values   
  
  2.  Inspection of missing values:   
Similar to our inspection of the 14x02 training data we are interested in knowing how many rows (if any) do not have any raw freezing data.  Again, we look for rows in which the value for all the variables containing the raw freezing data are "NA".     

First, we can look at a vector of the number of missing raw data values for each observation.   

```{r train_14x04 NA by row }

rowSums(is.na(train_14x04[,c(3,5,7,11,13,15)]))
```

From this vector, we can see that starting with row 98, there are several observations missing 6 of the 6 values.  We can now find how many observations are missing 6 values.  

```{r train_14x04 rows missing data}
## use select(), filter() and summarise() from dplyr

raw_dat <- train_14x04 %>%
    select(3,5,7,11,13,15) 

 num_NA_rows_14x04 <- raw_dat %>%
    filter(rowSums(is.na(raw_dat)) == 6) %>%
    summarise(num_rows = n())
 
 rm(raw_dat)
```

There are `r num_NA_rows_14x04[[1,1]]` observations without any raw freezing data.  These observations will be removed during the Data Cleaning step.  

```{r cleanup training vars, echo=FALSE, results='hide'}
rm(num_NA_rows_14x02, num_NA_rows_14x04)
```


## Training Data Cleaning   
In this step the identified issues in the 14x02 and 14x04 behavioral training data are addressed.  

### Select variables containing raw training data  
The dataset contains variables that contain values calculated from the raw data, rather than the raw data itself. These variables will be removed and later re-calculated during the Data Transformation step in order to maintain transparent and reproducible data analysis.   

```{r train data select vars}
## use the select() function of dplyr (tidyverse)

## select 14x02 vars
train_14x02 <- train_14x02 %>%
    select(1:3,5,7,11,13,15)

## select 14x04 vars
train_14x04 <- train_14x04 %>%
    select(1:3,5,7,11,13,15)
```

### Rename training data variables   
Some of the variable names are not in a correct syntax form.  All the variable names will be changed to more descriptive and syntaxically correct names.   

```{r rename train vars}
## set up a vector with the new variable names
vars <- c("rat", "training_day", "A_min1", "A_min2", "A_min3",
          "A'_min1", "A'_min2", "A'_min3")

## rename 14x02 vars
names(train_14x02) <- vars

## rename 14x04 vars
names(train_14x04) <- vars
```

We can inspect the resulting datasets:   
**14x02 training data: **   

```{r glimpse new 14x02 train dat}
glimpse(train_14x02)
```

**14x04 training data:**   

```{r glimpse new 14x04 train dat}
glimpse(train_14x04)
```

### Remove observations without training data   
The rows (observations) without any freezing data do not add to the dataset and will be removed.  For `train_14x02` any row missing 5 or 6 values will be removed, and for `train_14x04` any row missing 6 values will be removed.  

```{r rm NA rows from training data}
## remove rows missing data from train_14x02
train_14x02 <- train_14x02 %>%
    filter(!rowSums(is.na(train_14x02[,3:8])) >= 5)

## remove rows missing data from train_14x04
train_14x04 <- train_14x04 %>%
    filter(!rowSums(is.na(train_14x04[,3:8])) >= 5)
       
```

The `train_14x02` dataset now has `r dim(train_14x02)[1]` observations and the `train_14x04` dataset now has `r dim(train_14x04)[1]` observations   

### Add a 'study' variable to each training dataset
Here, we add a new variable for 'study' to each training dataset to distinguish the study as 14x02 or 14x04:   

```{r add study var to train}
## use the mutate() function of dplyr

## add the study value of 14x02
train_14x02 <- train_14x02 %>%
    mutate(study = "14x02")

## add the study value of 14x04
train_14x04 <- train_14x04 %>%
    mutate(study = "14x04")

```

### Combine 14x02 and 14x04 training datasets

Next, the rows of the 14x04 training data can be appended to the 14x02 training data to generate a compile `train_dat` data table.  The variables are then reordered.   

```{r append train 14x04 to 14x02}
## use the bind_rows() function of dplyr
train_dat <- bind_rows(train_14x02, train_14x04) 

## reorder the variables
train_dat <- select(train_dat,
                    1, 9, 2:8)
```

### Add a 'context_order' variable to the training data

Finally, we add context order information for each training day and reorder the variables:

```{r add context order}
## use the left_join() function from dplyr to add the context_order var and preserve the train_dat data
train_dat <- left_join(train_dat, context_order, by = "training_day")

## reorder the variables
train_dat <- select(train_dat,
                    1:3, 10, 4:9)
```

### Add the treatment 'group' variable to the training dataset   
One of the experimental objectives is to determine the effect of neuroinflammation on contextual memory.   Neuroinflammation was induced by systemic administration of lipopolysaccride (LPS).  Secondarily, to determine if LPS caused memory deficits via cytokine release by microglia, some rats were also treated with minocycline (MIN) to block microglia activation.  As a control, groups of rats were gives saline in place of either minocycline or LPS.  Thus, there were four treatment groups:   
  1.  saline-saline (SAL-SAL); control rats   
  2.  saline-LPS (SAL-LPS); rats given saline, instead of minocyline, followed by LPS   
  3.  minocycline-saline (MIN-saline); rats given minocycline, followed by saline   
  4.  minocycline-LPS (MIN-LPS); rats given minocycline, followed by LPS   
  5.  caged-control (CC); rats that not treated and are not tested in the context discrimination task   
  
Here the treatment group variable is added to the behavioral training data:   

```{r add grp to train dat}
## use the left_join() function from dplyr to add the group var and preserve the train_dat data
train_dat <- left_join(train_dat, grp_assigns, by = "rat")

## reorder the variables
train_dat <- select(train_dat,
                    1,11, 2:10)
```

### Convert training categorical variables to factor variables     

In the tidy training dataset (`train_dat`) there are `r dim(train_dat)[1]` observations and `r dim(train_dat)[2]` variables.  To facilitate downstream data analysis, the categorical variables 'study', 'group', 'training_day' and 'context_order' are converted to factor variables:   

```{r train_dat factor variables}
## use the mutate() function of dplyr
train_dat <- train_dat %>%
    mutate(study = as.factor(study),
           group = as.factor(group),
        training_day = as.factor(training_day),
        context_order = as.factor(context_order)
    )
```

Overview of final form of the behavioral training data:   

```{r final train_dat view}
glimpse(train_dat)
```

A sample of 20 observations from the `train_dat` dataset:   

```{r random rows train_dat}
## view random rows of the train_dat table
set.seed(563)

pander(head(sample_n(train_dat, 421), 20),
       caption = "Sample of the training data",
       caption.placement = "top",
       digits = 3, style = 'rmarkdown', split.table = "Inf")
```

```{r save train_dat table, echo=FALSE}
save(train_dat, file = "tidy_training.RData")
```

```{r cleanup 01, echo=FALSE, results='hide'}
save(train_14x02, file = "tidy_train_14x02.RData")
save(train_14x04, file = "tidy_train_14x04.RData")
save(train_dat, file = "tidy_training.RData")

rm(train_14x02, train_14x04, vars, context_order)
```

***   
## Initial Testing Data Inspection

Here we make an initial inspection of the imported behavioral testing datasets and identify issues to resolve in order to generate the corresponding tidy datasets.    

### 14x02 Testing Data Inspection   

  1.  Overview of the test_14x02 data:
  
```{r glimpse test_14x02}
## use glimpse() from the tibble package loaded with the tidyverse package

## overview of test_14x02
glimpse(test_14x02)
```
  
From this view of the 14x02 testing data we see that:   
  *  There are `r dim(test_14x02)[1]` rows and `r dim(test_14x02)[2]` variables   
  *  The data appear to be in a long, tidy format   
  *  Rat identifiers are stored in the "Subject" variable  
  *  Rat identifiers "14x02_1", "14x02_2", "14x02_3", "14x02_5", "14x02_7", "14x02_9", should be in the form of "14x02_0#"   
  *  The raw freezing data of interest is in numeric variables with names containing the phrases "Min n sec".  There are 5 columns with raw context A freezing data and 3 columns with raw context A' freezing data   
  *  The remaining numeric variables are values calculated from the raw data 
  *  Freezing data for context A is stored in variables with names starting with "S+"   
  *  Freezing data for context A' is stored in variables with names starting with "S-"   
  
  2.  Inspection of missing values:   
We will use the same procedure here as we used for the training data to determine if- and how many rows do not contain raw freezing data.       

First, we can look at a vector of the number of missing raw data values for each observation.   

```{r test_14x02 NA by row 01}
rowSums(is.na(test_14x02[,c(2:6, 14:16)]))
```

From this vector, we can see that observation 18 has 1 missing value.  We can take a closer look at this row.    
```{r test_14x02row18}
pander(test_14x02[18,c(1:6, 14:16)])
```

We can see that the only value missing is that for minute 2 of context A'.  Since, the other values for contex A' are 0, we will assume that the value of minute 2 is also 0 and will set the value to 0.   

```{r test_14x02 NA to 0 }
## change the NA in row 18, column 15 to 0
test_14x02[18,15] <- 0
```

Another look:   

```{r test_14x02 NA by row 02}
rowSums(is.na(test_14x02[,c(2:6, 14:16)]))
```

Thus, the test_14x02 dataset does not have any missing values for the raw freezing data.   

### 14x04 Testing Data Inspection   

  1.  Overview of the `test_14x04` dataset:
  
```{r glimpse test_14x04}
## use glimpse() from the tibble package loaded with the tidyverse package

## overview of test_14x04
glimpse(test_14x04)
```
  
From this view of the 14x04 testing data we can see that:   
  *  There are `r dim(test_14x04)[1]` rows and `r dim(test_14x04)[2]` variables   
  *  The data appears to be in exactly the same format as the `test_14x02` dataset   
  *  The variables are exactly the same as in the `test_14x02` dataset  
  *  The dataset contains rats from both 14x04 **and** 14x02 (which will be removed)
  
  2.  Inspection of missing values:   
We will use the same procedure here as we used for the `test_14x02` dataset to determine if- and how many rows do not contain raw freezing data.       

We can look at a vector of the number of missing raw data values for each observation.   

```{r test_14x04 NA by row}
rowSums(is.na(test_14x04[,c(2:6, 14:16)]))
```

Row 1 of the `test_14x04` dataset has 1 missing value.  However, from the overview of the dataset, we can see that the first row is for a rat in the 14x02 study.  This row will be removed during the data cleaning step and we don't need to do anything here.  

## Testing data Data Cleaning   
In this step the identified issues in the 14x02 and 14x04 behavioral testing data are addressed.  

### Select variables containing raw testing data  
The dataset contains variables that contain values calculated from the raw data, rather than the raw data itself. These variables will be removed and later re-calculated during the Data Transformation step in order to maintain transparent and reproducible data analysis.   

```{r test data select vars}
## use base subsetting to get around problem of duplicate var name
## select 14x02 vars
test_14x02 <- test_14x02[, c(1:6,14:16)]
    
## select 14x04 vars
test_14x04 <- test_14x04[, c(1:6,14:16)]
```

### Rename testing data variables   
Some of the variable names are not in a correct syntax form.  All the variable names will be changed to more descriptive and syntaxically correct names.   

```{r rename testing vars}
## set up a vectors with the new variable names
vars <- c("rat", "A_min1", "A_min2", "A_min3",
          "A_min4", "A_min5",
          "A'_min1", "A'_min2", "A'_min3")


## rename 14x02 vars
names(test_14x02) <- vars

## rename 14x04 vars
names(test_14x04) <- vars
```

### Correct inconsistent rat identifiers in 14x02 testing data   
Rat identifiers "14x02_1", "14x02_2", "14x02_3", "14x02_5", "14x02_7", "14x02_9", should be in the form of "14x02_0#".   

```{r correct 14x02 rat ids}

test_14x02$rat <- ifelse(test_14x02$rat == "14x02_1", "14x02_01",
                    ifelse(test_14x02$rat == "14x02_2", "14x02_02",
                    ifelse(test_14x02$rat == "14x02_3", "14x02_03",
                    ifelse(test_14x02$rat == "14x02_5", "14x02_05",
                    ifelse(test_14x02$rat == "14x02_7", "14x02_07",
                    ifelse(test_14x02$rat == "14x02_9", "14x02_09",
                           test_14x02$rat
                           ))))))
```


We can inspect the resulting datasets:   
**14x02 testing data: **   

```{r glimpse new 14x02 test dat}
glimpse(test_14x02)
```

**14x04 testing data:**   

```{r glimpse new 14x04 test dat}
glimpse(test_14x04)
```

### Remove 14x02 observations from the 14x04 testing dataset   
During inspection of the 14x04 testing data it was found that the dataset also contained some 14x02 subjects.  These rats are removed from the 14x04 dataset here.   

```{r rm 14x02 rats from 14x04}
## use the str_detect() function from the stringr package
test_14x04 <- test_14x04 %>%
    filter(str_detect(rat, "14x04"))
```

### Add the 'study' variable to testing data   
Here, we add a new variable for 'study' to each testing dataset to distinguish the study as 14x02 or 14x04:   

```{r add study var to test}
## use the mutate() function of dplyr

## add the study value of 14x02
test_14x02 <- test_14x02 %>%
    mutate(study = "14x02")

## add the study value of 14x04
test_14x04 <- test_14x04 %>%
    mutate(study = "14x04")

```

### Combine 14x02 and 14x04 training datasets

Next, the rows of the 14x04 testing data are appended to the 14x02 testing data to generate a compile `test_dat` data table.  The variables are then reordered.   

```{r append test 14x04 to 14x02}
## use the bind_rows() function of dplyr
test_dat <- bind_rows(test_14x02, test_14x04) 

## reorder the variables
test_dat <- select(test_dat,
                    1, 10, 2:9)
```

```{r save test dats, echo=FALSE}
save(test_14x02, file = "tidy_test_14x02.RData")
save(test_14x04, file = "tidy_test_14x04.RData")

rm(vars, test_14x02, test_14x04)
```

### Add the treatment 'group' variable to the testing dataset   
Here the treatment group variable is added to the behavioral testing data:   

```{r add grp to test dat}
## use the left_join() function from dplyr to add the group var and preserve the test_dat data
test_dat <- left_join(test_dat, grp_assigns, by = "rat")

## reorder the variables
test_dat <- select(test_dat,
                    1,11, 2:10)
```

### Convert testing categorical variables to factor variables     

In the tidy testing dataset (`test_dat`) there are `r dim(test_dat)[1]` observations and `r dim(test_dat)[2]` variables.  To facilitate downstream data analysis, the categorical variables 'study', and 'group' are converted to factor variables:   

```{r test_dat factor variables}
## use the mutate() function of dplyr
test_dat <- test_dat %>%
    mutate(study = as.factor(study),
           group = as.factor(group))
```


Overview of final form of the behavioral testing data:   

```{r final test_dat view}
glimpse(test_dat)
```

A sample of 20 observations from the `test_dat` dataset:   

```{r random rows train_dat}
## view random rows of the test_dat table
set.seed(543)

pander(head(sample_n(test_dat, 42), 20),
       caption = "Sample of the testing data",
       caption.placement = "top",
       digits = 3, style = 'rmarkdown', split.table = "Inf")
```

```{r save test_dat table, echo=FALSE}
save(test_dat, file = "tidy_testing.RData")
```

```{r rm grp_assigns, echo=FALSE, results='hide'}
rm(grp_assigns)
```


## R session information:
```{r session info, echo=FALSE}
devtools::session_info(
    c("tidyverse", "readxl", "broom", "stringr",
    "rmarkdown", "knitr", "pander"))

```
