---
title: "Inspection of the discrimination ratio"
author: "Gail Lewandowski"
date: "February 28, 2017"
output: github_document    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, echo=FALSE, results='hide'}
suppressMessages({
    library(tidyverse)
    library(rmarkdown)
    library(knitr)
})
```

## Discrimination Ratio   
Contextual discrimination memory was assessed by measuring freezing in two similar behavioral chambers.  There was a mild footshock in context A but not in context A'.  Rats were were trained to criterion as defined by >=25% freezing in A **and** <=10% freezing in A' for **2 consecutive days**.  Memory strength (or sensitivity) is proportional to the % freezing in context A, and memory specificity is inversely proportional to the % freezing in context A'.  The data from both contexts are often presented together; freezing in A increases as freezing in A' decreases.   
Another measure used is the **Discrimination Ratio** (DR) which is defined as:   
    %A-%A' / %A+%A'   

Here the validity of the DR as a stand-alone measure of context discrimination conditioning (CDC) is examined using the training data from 42 rats.   

```{r load training_dat table}
load(file = "training_data.RData")
```

### Addition of the summary DR statistic to the training data

In this section a copy of the `training_dat` table will be made (`training_dat2`) and the summary variable `DR` (discrimination ratio) will be calculated and added to it.  

```{r add DR variable}
## add the DR variable to a copy of training_dat
training_dat2 <- training_dat %>%
    mutate(DR = (training_A - `training_A'`)/
               (training_A + `training_A'`))

## round to 3 decimal places
training_dat2[,"DR"] <- round(training_dat2[,"DR"],3)
```

A rat has a successful context discrimination session if freezing in A is >=25 **AND** freezing in A' is <=10.  Here the variable `CDC` (context discrimination conditioning) is added to indicate context discrimination (CD) or non-discrimination (ND).   

```{r add D variable}
## condition for sucessful CDC event
con <- training_dat2$training_A >= 25 & training_dat2$`training_A'`<=10 

## Add the CD variable
training_dat2$CDC <- ifelse(con, "CD", "ND")

remove(con)
```

Generate a new table that has just the final day of training for each rat.   

```{r final training day}
## subset training_dat2 for the last training day for each rat
final_training_day <- training_dat2 %>%
    group_by(rat) %>%
    filter(training_day == max(training_day))

## inspect the table
head(final_training_day)
```

### Exploratory analysis of the discrimination ratio   

Exploratory graphs can be used to investigate the relationship between the discrimination ratio and successful context discrimination events.  

#### Exploratory graphs of all training data   
    1. Distribution of DR for all rats by day   
    
```{r all trainingdata all DR}
## 
training_dat2 %>% 
    filter(is.na(DR)==FALSE) %>%
    ggplot(aes(x= training_day, 
                       y= DR, color = CDC)) +
    geom_point(position = "jitter") +
    geom_smooth(na.rm = TRUE, fullrange = TRUE, 
                span = 0.8, se = 0)

training_dat2 %>% 
    filter(is.na(DR)==FALSE) %>%
    ggplot(aes(x=DR, fill = CDC) ) + 
    geom_density(alpha=.3, na.rm = TRUE) +
    facet_wrap(~ training_day, nrow = 5)


dat <- training_dat2 %>%
    group_by(training_day) %>%
    summarise(n_rats = n_distinct(rat),
              n_cd = sum(CDC == "CD"), 
              n_nd = sum(CDC == "ND")) %>%
    ggplot(aes(x = training_day))+
    geom_point(mapping = aes(y= n_rats),color = "blue", size =2) +
    geom_smooth(mapping = aes(y= n_rats),
                color = "blue", se = 0) +
    geom_point(mapping = aes(y= n_cd),color = "green", size =2) +
    geom_smooth(mapping = aes(y= n_cd),
                color = "green", se = 0) +
    geom_point(mapping = aes(y= n_nd),color = "red", size =2) +
    geom_smooth(mapping = aes(y= n_nd),
                color = "red", se = 0)
    


    
p1 <- numCD %>%
    select(training_day, n_nd) %>%
    ggplot(aes(x = training_day, y = n_nd )) +
    geom_point(color = "blue", size = 2)
    
    
    
    




```

    2. Distribution of DR >= 0.5 for all rats by day  
    
```{r all training data DR>=0.5}

    
    
    ggplot(aes(x= training_day, 
                       y= DR, color = CDC)) +
    geom_point(position = "jitter") +
    geom_smooth(na.rm = TRUE, fullrange = TRUE, 
                span = 0.8, se = 0)

training_dat2 %>% 
    filter(is.na(DR)==FALSE,
           DR >= 0.5) %>%
    ggplot(aes(x=DR, fill = CDC) ) + 
    geom_density(alpha=.3, na.rm = TRUE) +
    facet_wrap(~ training_day, nrow = 5)
```


Exploratory graphs:  

```{r}
ggplot(training_dat2, aes(x=DR, fill=CD)) + 
    geom_density(alpha=.3, na.rm = TRUE, ylim = 5.0) +
    facet_wrap(~ training_day, nrow = 5)

ggplot(data = training_dat2, aes(x= training_day, 
                       y= DR, color = CD)) +
    geom_point(na.rm = TRUE, position = "jitter") +
    geom_smooth(na.rm = TRUE, 
                fullrange = FALSE, 
                span = 0.8, se = 0) +
    facet_wrap(~ Group, nrow = 2)



ggplot(data = training_dat2, aes(x = DR, fill=CD)) +
    geom_histogram(alpha = 0.3, na.rm = TRUE, 
                   binwidth = 0.5, position = "dodge")

#############################################################

## distribution of freezing in A
### All training data
ggplot(training_dat2, aes(x=training_A)) + 
    geom_density(alpha=.3, na.rm = TRUE, fill = "blue")

## by CD vs ND
ggplot(training_dat2, aes(x=training_A, fill=D)) + 
    geom_density(alpha=.3, na.rm = TRUE)

ggplot(training_dat2, aes(x=training_A, y = DR, color=D)) + 
    geom_point(alpha=.3, na.rm = TRUE)

## final training day
### All final day data
ggplot(final_training_day, aes(x=training_A)) + 
    geom_density(alpha=.3, na.rm = TRUE, fill = "blue")

## by CD vs ND
ggplot(final_training_day, aes(x=training_A, fill=D)) + 
    geom_density(alpha=.3, na.rm = TRUE)

ggplot(final_training_day, aes(x=training_A, y = DR, color=D)) + 
    geom_point(alpha=.3, na.rm = TRUE)






```













Here the validity of the DR as a measure of context discrimination conditioning (CDC) using values between 0 and 100 randomly generated for A and A'

```{r ficticious CDC data}
set.seed(3456)
x <- 0:90

A1 <- sample(x, 500, replace = TRUE)
A2 <- sample(x, 500, replace = TRUE)

fict_dat <- tbl_df(data.frame(A1,A2))
fict_dat <- fict_dat %>%
    mutate(DR = (A1-A2)/(A1+A2))

## round to 3 decimal places
fict_dat[,"DR"] <- round(fict_dat[,"DR"],3)

## add D variable
con <- fict_dat$A1 >= 25 & fict_dat$A2 <=10 
fict_dat$D <- ifelse(con, "CD", "ND")

ggplot(fict_dat, aes(x=DR, fill=D)) + 
    geom_density(alpha=.3, na.rm = TRUE)

ggplot(data = training_dat2, aes(x = DR, fill=D)) +
    geom_histogram(alpha = 0.3, na.rm = TRUE, 
                   binwidth = 0.5, position = "dodge")

```



```{r exploratory graphs}

ggplot(fict_dat, aes(x=DR, fill=D)) + geom_density(alpha=.3)


ggplot(data = fict_dat, aes(x= DR)) +
    geom_density(mapping = aes(color = D,
                               fill = D),
                 alpha = 0.5,             
                 na.rm = TRUE)


```










Here the validity of the DR as a measure of context discrimination conditioning (CDC) is examined using the training data from 42 rats.   


```{r load training_dat table}
load(file = "training_data.RData")
```

In this section the summary variable `DR` will be calculated and added to the `training_dat` data table.  

```{r add DR variable}
## add the DR variable to a copy of training_dat
training_dat2 <- training_dat %>%
    mutate(DR = (training_A - `training_A'`)/
               (training_A + `training_A'`))

## round to 3 decimal places
training_dat2[,"DR"] <- round(training_dat2[,"DR"],3)
```

A rat has a successful context discrimination session if freezing in A is >=25 **AND** freezing in A' is <=10.  Here the variable `D` is added to indicate context discrimination (1) or non-discrimination (0).   

```{r add D variable}
con <- training_dat2$training_A >= 25 & training_dat2$`training_A'`<=10 

training_dat2$D <- ifelse(con, "CD", "ND")

```

Generate a new table that has just the final day of training for each rat.   

```{r final training day}
final_training_day <- training_dat2 %>%
    group_by(rat) %>%
    filter(training_day == max(training_day))
```

















































